A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-07-11 example.tex
explanation

\begin{example}\label{ex:string-constraints} Consider the following set of
    string constraints over string variables~$s_1, s_2$ and integer
    variables~$n, i$:
\begin{constraints}
    \item\label{const:s1-in-c-dd} $s_1 \in \Language(\Regex{c\RegexOr{}dd})$.
    \item\label{const:s2-in-b} $s_2 \in \Language(B)$, the language accepted by
    automaton~$B$ of \cref{fig:aut_b}.
    \item\label{const:s1-substring} $s_1 = \Substring(s_2, i, n)$, that is $s_1$ is an
    $n$-length substring of $s_2$ starting at offset $i$.
    \item\label{const:more-inside-than-before} $n > i$, that is what comes
    before $s_1$ in $s_2$ is shorter than $s_1$.
    \item\label{const:something-before-and-after} $i > 0 \land \Length{s_2} -i -n > 0$, that
    is there is at least one character in $s_2$ before and after $s_1$.
\end{constraints}
\end{example}

Although the constraints are simple, it should be noted that
decidability results for string constraints involving integers are
notoriously hard to obtain; most state-of-the-art string solvers, for
instance Z3~\cite{Z3}, different versions of
Z3-str~\cite{Z3-str,DBLP:conf/fm/MoraBKNG21}, or
cvc5~\cite{cvc5}
are not guaranteed to be complete or to terminate on constraints of this kind.

A decision procedure for constraints of this kind is defined in
\cite{ostrich-plus}.
%
To solve the constraints, we will first perform translation into Parikh
automata, then break down how the constraints will be handled by \Calculus{}.
This example will recur in a more formal fashion in \cref{sec:generalised},
which can be read in parallel if (even) more detail is desired.

@@

To solve \cref{ex:string-constraints} using \Calculus{} in a theorem proving
context, \cref{const:s1-in-c-dd,const:s2-in-b,const:s1-substring} can be
translated to the product of Parikh automata $B \times A'$, where $A'$
(\cref{fig:aut_a}) is the Parikh automata obtained by applying the encoding of
$\Substring$ described in \cite{ostrich-plus} to
$\Language(\Regex{c\RegexOr{}dd})$. That description can then be used to obtain
values for the substring offsets $i, n$. The corresponding automata can be found
in \cref{fig:examples}.

In the automata of \cref{fig:examples}, we will use the notation $a /
\begin{bmatrix} x_+ \end{bmatrix}$ to mean that a transition reads an input
character $a$ and increments register $x$ by one. We will omit zero-valued
increments and assume that all registers are scoped to their automaton. The
increments are usually represented as a vector (hence the brackets), but as it
is mostly sparse here, we use the symbolic notation $\begin{bmatrix} x_{1+}
\end{bmatrix}$ rather than the more cumbersome $\left[ 0, 1,  0 , \ldots, 0
\right]$.

@@

Intuitively, \cref{ex:string-constraints} are unsatisfiable. By inspecting the
automata, we realise that the path using $\Char{dd}$ in $A'$ is unusable since
are no $\Char{d}$-labelled transitions in $B$. This means that $s_1 = \Char{c}$,
and thus $n = 1$. That means that \cref{const:more-inside-than-before} implies
that $i = 0$. However, no path through $B$ passes by a $\Char{c}$ without a
preceding character. Therefore, already
\cref{const:s1-in-c-dd,const:s1-in-c-dd,const:s1-substring,const:more-inside-than-before}
together are unsatisfiable.

We will now proceed to show how we prove this using the three principles of
\Calculus{}: problem-aware case splitting (sometimes called branching), lazy
enforcement of connectivity constraints (essentially consisting of a constraint
programmgin-style propagator), and lazy computation of products of automata.

>>>>

\begin{example}\label{ex:string-constraints} Consider the following set of
    string constraints over string variables~$s_1, s_2$ and integer
    variables~$n, i$:
\begin{constraints}
    \item\label{const:s1-in-c-dd} $s_1 \in \Language(\Regex{c\RegexOr{}dd})$.
    \item\label{const:s2-in-b} $s_2 \in \Language(B)$, the language accepted by
    automaton~$B$ of \cref{fig:aut_b} (ignoring
    the counter~$l_B$).
    \item\label{const:s1-substring} $s_1 = \Substring(s_2, i, n)$, that is $s_1$ is an
    $n$-length substring of $s_2$ starting at offset $i$.
    \item\label{const:more-inside-than-before} $n > i$, that is what comes
    before $s_1$ in $s_2$ is shorter than $s_1$.
    \item\label{const:something-before-and-after} $i > 0 \land \Length{s_2} -i -n > 0$, that
    is there is at least one character in $s_2$ before and after $s_1$.
\end{constraints}
\end{example}

Although the constraints are simple, it should be noted that
decidability results for string constraints involving integers are
notoriously hard to obtain; many state-of-the-art string solvers, for
instance Z3~\cite{Z3}, different versions of
Z3-str~\cite{Z3-str,DBLP:conf/fm/MoraBKNG21}, or
cvc5~\cite{cvc5}
are not guaranteed to be complete or to terminate on such constraints
(but, of course, they often achieve very good performance in practice).

A decision procedure for a fragment of string constraints with integers,
covering our constraints, is defined in
\cite{ostrich-plus}.
%
We will first perform translation into Parikh
automata, then break down how the constraints will be handled by \Calculus{}.
This example will recur in a more formal fashion in \cref{sec:generalised},
which can be read in parallel if (even) more detail is desired.

@@

To solve \cref{ex:string-constraints} using \Calculus{} in a theorem
proving context,
\cref{const:s1-in-c-dd,const:s2-in-b,const:s1-substring} can be
translated to the product of Parikh automata $B \times
A'$. Automaton~$B$ is given in \cref{fig:aut_b}.  Automaton~$A'$
(\cref{fig:aut_a}) is an automaton defining the pre-image of
$\Language(\Regex{c\RegexOr{}dd})$ under the $\Substring$ function,
and obtained by applying the construction described in
\cite{ostrich-plus}. Intuitively, $A'$ describes all
tuples~$(s_2, i, n)$ such that $\Substring(s_2, i, n)$ is in
$\Language(\Regex{c\RegexOr{}dd})$. To this end, $A'$ contains
registers~$l_A, i, n$, which count the length of the overall string
read, the length of the prefix before the extracted substring, and the
length of the substring, respectively. Registers are assumed to start
from zero in each automaton run.

In the automata of \cref{fig:examples}, we use the notation $a /
\begin{bmatrix} x_+ \end{bmatrix}$ to mean that a transition reads an input
character $a$ and increments register $x$ by one. We will omit zero-valued
increments and assume that all registers are scoped to their automaton. The
increments are usually represented as a vector (hence the brackets), but as it
is mostly sparse here, we use the symbolic notation $\begin{bmatrix} x_{1+}
\end{bmatrix}$ rather than the more cumbersome $\left[ 0, 1,  0 , \ldots, 0
\right]$.

@@

Intuitively, \cref{ex:string-constraints} is unsatisfiable. By inspecting the
automata, we realise that the path using $\Char{dd}$ in $A'$ is unusable since
are no $\Char{d}$-labelled transitions in $B$. This means that $s_1 = \Char{c}$,
and thus $n = 1$. That means that \cref{const:more-inside-than-before} implies
that $i = 0$. However, no path through $B$ passes by a $\Char{c}$ without a
preceding character. Therefore, already
\cref{const:s1-in-c-dd,const:s1-in-c-dd,const:s1-substring,const:more-inside-than-before}
together are unsatisfiable.

We will now proceed to show how we prove this using the three principles of
\Calculus{}: problem-aware case splitting (sometimes called branching), lazy
enforcement of connectivity constraints (essentially consisting of a constraint
programming-style propagator), and lazy computation of products of automata.
